Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-05-23 TEX/draft4.tex
Afternoon session

\TODO{moments approach + announce the results}

\TODO{Paragraph about bias; can also mention computational aspects (single pass, etc.)}

@@

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of data sequences (micrographs) of size $250\times 250$ with additive i.i.d.\ Gaussian noise with variance $\sigma^2$. Each micrograph contains four occurrences of a $50 \times 50$ image of Einstein. Figure~\ref{fig:micro_example} shows a simple example of our problem with  one signal ($K=1$). 
		Each panel demonstrates an image of size $250\times 250$, containing 4 occurrences of an Einstein image of size $50\times 50$. 
		The left panel shows the data without noise. In the middle panel,  i.i.d.\ Gaussian noise with standard deviation $\sigma=0.5$  was added, however,  Einstein is still easily identified. In this scenario it is likely that standard detection algorithms can locate the copies of the signal within the micrograph. In the right panel, we show the same data swamped in  
		i.i.d.\ Gaussian noise with standard deviation  $\sigma=3$.
		Clearly, identifying the locations of Einstein in this micrograph is much more challenging. In fact, it can be shown that in the low signal--to--noise ($\SNR$) regime, 
		detection  of individual image occurrences is impossible---even if the true image is known---and therefore particle picking is impossible.  This phenomenon is explained in detail in Appendix~\ref{sec:theory}.}	
\end{figure}

\paragraph{Autocorrelation analysis.}
We now return to the general problem of estimating $x_1,\dots,x_K$ from the noisy observation $y$.
In order to recover the signals $x_1,\dots,x_K$ from the data, we use autocorrelation analysis.
In a nutshell, the method consists of two stages. First, we estimate a mixture (i.e., linear combination) of the low--order autocorrelation functions of the signals from the data. These quantities can be estimated, to any desired accuracy, if individual occurrences are separated by the maximum signal size and each signal appears sufficiently many times in the data. There is no need to detect individual occurrences.
In the second stage, the signals are estimated from the mixed autocorrelations using a nonconvex least-squares (LS) algorithm or a phase retrieval algorithm; see Section~\ref{sec:methods} for details.
This method requires only one pass over the data and can recover the signals in any $\SNR$, if the signals appear enough times in the data. As a side note, we mention that expectation-maximization---a popular framework in SPR---is intractable for this problem (see Appendix~\ref{sec:theory}  for more details).

>>>>

In order to recover the planted image, we use autocorrelation analysis. In a nutshell, we relate the autocorrelation functions of the micrographs to the autocorrelation functions of the planted image.
%To do so, we assume the image occurrences are well separated.
For any noise level, these autocorrelations can be estimated to any desired accuracy provided individual occurrences of the image are well separated and the image appears sufficiently many times in the micrographs. Importantly, there is no need to detect individual occurrences. The autocorrelations of the micrographs are straightforward to compute and require only one pass over the data. These directly yield estimates for the autocorrelations of the image. To estimate the image itself from its estimated autocorrelations, we solve a nonlinear inverse problem via least-squares or a phase retrieval algorithm; see Figure~\ref{fig:Einst_example} for an illustration and Section~\ref{sec:methods} for details. As a side note, we mention that expectation-maximization (EM)---a popular framework in SPR---is intractable for this problem; see Appendix~\ref{sec:theory} for a discussion.

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{shatsky2009method,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using cross-correlation and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates are required.

\TODO{What about this one? \cite{vanheel1992correlation}}

%\TODO{Paragraph about bias; can also mention computational aspects (single pass, etc.)}

@@

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contains four occurrences of the same $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low $\SNR$, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}	
\end{figure}

%\paragraph{Autocorrelation analysis.}
%We now return to the general problem of estimating $x_1,\dots,x_K$ from the noisy observation $y$.
%In order to recover the signals $x_1,\dots,x_K$ from the data, we use autocorrelation analysis.
%In a nutshell, the method consists of two stages. First, we estimate a mixture (i.e., linear combination) of the low--order autocorrelation functions of the signals from the data. These quantities can be estimated, to any desired accuracy, if individual occurrences are separated by the maximum signal size and each signal appears sufficiently many times in the data. There is no need to detect individual occurrences.
%In the second stage, the signals are estimated from the mixed autocorrelations using a nonconvex least-squares (LS) algorithm or a phase retrieval algorithm; see Section~\ref{sec:methods} for details.
%This method requires only one pass over the data and can recover the signals in any $\SNR$, if the signals appear enough times in the data. As a side note, we mention that expectation-maximization---a popular framework in SPR---is intractable for this problem (see Appendix~\ref{sec:theory}  for more details).
